---
title: "Practical Machine Learning Course Project"
author: "Donson Yeh"
date: "2017.9.2"
output:
  html_document: default
  pdf_document: default
---


##Background and Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ¡V a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
In this project, we will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participant They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The five ways are exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Only Class A corresponds to correct performance. The goal of this project is to predict the manner in which they did the exercise, i.e., Class A to E. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data Processing
#Import the data
We will load the necessarry packages for analyzing first and then import the training and testing data set from the given URLs.

```{r}
# load the required packages
library(caret)
library(rattle)
library(rpart)
library(rpart.plot)
library(randomForest)
library(repmis)
# import the data from the URLs
trainurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training <- source_data(trainurl, na.strings = c("NA", "#DIV/0!", ""), header = TRUE)
testing <- source_data(testurl, na.strings = c("NA", "#DIV/0!", ""), header = TRUE)
# load data locally
training <- read.csv("pml-training.csv", na.strings = c("NA", ""))
testing <- read.csv("pml-testing.csv", na.strings = c("NA", ""))
```

The training dataset has 19622 observations and 160 variables, and the testing data set contains 20 observations and the same variables as the training set. We are trying to predict the outcome of the variable classe of the training set.

#Data cleaning
We will remove some columns (predictors) of the training set which contain missing values.

```{r}
training <- training[, colSums(is.na(training)) == 0]
testing <- testing[, colSums(is.na(testing)) == 0]
```

We also remove the first seven predictors since these variables have little predicting contribution for the outcome classe.

```{r}
trainData <- training[, -c(1:7)]
testData <- testing[, -c(1:7)]
```

Then we have the cleaned data sets trainData and testData which have 53 columns with the same 52 variables and the last variable classe andproblem_id individually. The trainData has 19622 rows while the testData has 20 rows.

#Data spliting
For the sake of getting the out-of-sample errors, we will split the cleaned training set trainData into a training set (70%) for prediction and a validation set (30%) to calculate the out-of-sample errors.

```{r}
set.seed(7826) 
inTrain <- createDataPartition(trainData$classe, p = 0.7, list = FALSE)
train <- trainData[inTrain, ]
valid <- trainData[-inTrain, ]
```

##Prediction Algorithms
We will try the classification trees and random forests to predict the outcome.

#Classification trees
We will use 5-fold cross validation to implement the algorithm. Since the data transformations may be less important in non-linear models like classification trees, we will not transform any variables.

```{r}
control <- trainControl(method = "cv", number = 5)
fit_rpart <- train(classe ~ ., data = train, method = "rpart", trControl = control)
print(fit_rpart, digits = 4)
fancyRpartPlot(fit_rpart$finalModel)
```

```{r}
# predict outcomes using validation set
predict_rpart <- predict(fit_rpart, valid)
# Show prediction result
(conf_rpart <- confusionMatrix(valid$classe, predict_rpart))
(accuracy_rpart <- conf_rpart$overall[1])
```

From the confusion matrix, the accuracy rate is 0.5, and so the out-of-sample error rate is 0.5. It seems that the classification tree doesn't predict very well.

#Random forests
Since classification tree method doesn't predict very well, we try the random forest algorithm instead.

```{r}
fit_rf <- train(classe ~ ., data = train, method = "rf", trControl = control)
print(fit_rf, digits = 4)
# predict outcomes using validation set
predict_rf <- predict(fit_rf, valid)
# Show prediction result
(conf_rf <- confusionMatrix(valid$classe, predict_rf))
(accuracy_rf <- conf_rf$overall[1])
```

The random forest algorithm perform better than the classification tree. The accuracy rate is up to 0.991, and so the out-of-sample error rate is 0.009. This may be due to the fact that many predictors are highly correlated. Random forests chooses a subset of predictors at each split and decorrelate the trees. This leads to high accuracy, although this algorithm is sometimes difficult to interpret.

#Prediction on Testing Set
We choose the random forests algorthm as the prediction of the outcome variable classe of the testing set.

```{r}
(predict(fit_rf, testData))
```

We can predict new data into a classe by the radom forest algorithm.